---
title: "ST661 Project"
author: "Sahil Shashin Shah"
output:
  html_document:
    code_folding: hide
header-includes:
 \usepackage{float}
 \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Weather in Ireland in 2017 {.tabset .tabset-pills}

## Introduction {.tabset .tabset-pills}

### Dataset Details {.tabset .tabset-pills}

This is the aimsir17 dataset, it is from the aimsir17 library, it contains 
details on the minutely weather in Ireland in 2017. It has three subsets: the 
eirgrid17 data, the observations data and data on the stations. 

The eirgrid17 set contains 35,040 rows of data with fifteen columns: year, 
month, day, hour, minute, date, NIGeneration, NIDemand, NIWindAvailability,
NIWindGeneration, IEGeneration, IEDemand, IEWindAvailability, IEWindGeneration 
and SNSP. The IE refers to the Republic of Ireland while the NI 
refers to Northern Ireland, consisting of the counties; Antrim, Armagh, Down, 
Fermanagh, Derry and Tyrone. 

The observations set contains 219,000 rows of data with twelve columns: station,
year, month, day, hour, date, rain, temp, rhum, msl, wdsp and wddir.

The stations set contains 25 rows of data with 5 columns: station, county, 
height, latitude and longitude.

### Column Explanations {.tabset .tabset-pills}

The following are explanations for the columns being used in the analysis.

Rain: rainfall in mm.

Temp: temperature in degrees Celsius.

Rhum: relative humidity percentage.

Msl: mean sea level pressure in hPa.

Wdsp: wind speed in knots, later converted to km/hr.

Wddir: wind direction in degrees.

### We will Discuss {.tabset .tabset-pills}

We will be analysing how the different variables in the observations data affect 
each other, as well as their trends throughout the year.

We will then focus on wind speed and mean sea level pressure exclusively from 
the 16th of October, which is when storm Ophelia occurred.

We will also look at the data from the day the storm occurred.

Then, we will remove the 16th of October from the data set and compare it 
with the original data to see how much it affected the country's total wind 
speed for the year.

### Methods {.tabset .tabset-pills}

We will be using four libraries: tidyverse, ggplot2, gridExtra and ggpubr

## Data Cleaning {.tabset .tabset-pills}

```{r}
library(aimsir17)
eirgrid <- aimsir17::eirgrid17
observations <- aimsir17::observations
stations <- aimsir17::stations
```

```{r include  = FALSE}
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(ggpubr)
```

Using the merge() function from the tidyverse package, the eirgird17 and 
observations data were joined, then that join was merged with the stations data. 
Then the rows that contained NA values were omitted using na.omit(). The 
following columns were also omitted; the minutes, SNSP, NIGeneration, NIDemand, 
NIWindAvailability, NIWindGeneration, IEGeneration, IEDemand, IEWindAvailability
and IEWindGeneration as we will not be using them.
```{r results = FALSE}
join <- merge(eirgrid, observations)
join2 <- (merge(join,stations))
colnames(join2)[colSums(is.na(join2)) > 0]
nrow(join2)
data <- na.omit(join2)
data <- data[ , !(names(data) %in% c("minute","SNSP","NIGeneration","NIDemand",
                                     "NIWindAvailability","NIWindGeneration",
                                     "IEGeneration","IEDemand",
                                     "IEWindAvailability","IEWindGeneration"))]
```

```{r}
data$date <- sub(" .*", "", data$date)
data$wdsp <- (data$wdsp*1.852) #converting wdsp from knots to km/hr
```

```{r}
data$month <- factor(data$month)
levels(data$month) <- c("January","February","March","April","May",
                                 "June","July","August","September","October",
                                 "November","December")
```

```{r}
october <- data %>%
  filter(data$month == "October")
 
october_df <- data.frame(october)
```

```{r}
storm_ophelia <- data %>%
  filter(date =="2017-10-16")
```

```{r}
not_storm_ophelia <- data %>%
  filter(date != "2017-10-16")
```

```{r}
year <- data$year
month <- data$month
day <- data$day
hour <- data$hour
date <- data$date
station <- data$station
rain <- data$rain
temp <- data$temp
rhum <- data$rhum
msl <- data$msl
wdsp <- data$wdsp
wddir <- data$wddir
county <- data$county
height <- data$height
latitude <- data$latitude
longitude <- data$longitude
```

There were NA values in rain, temp, rhum, msl, wdsp, wddir columns.
So, in total, 17,757 rows were omitted from the data.

There were no inappropriate data variable types.

## Analysis {.tabset .tabset-pills}

### How the Variables Affect Each Other {.tabset .tabset-pills}

```{r, fig.cap = "Figure 1.1"}
pairs(rain ~ wdsp + wddir + temp + rhum + msl, data, 
      col = "green", 
      pch = 18, 
      labels = c("Rain", "Wind Speed", "Wind Direction", "Temperature", 
                 "Humidity", "Pressure"),
      main = "Pairs Plot of the Weather Variables")
```
Figure 1.1 is an overall paired plot of factors involved in the 
observation table. As the mean sea level pressure increases, the rainfall in mm 
decreases. When there is more rainfall, humidity is also high. As the mean sea 
level pressure decreases, the wind speed increases.

### Finding Storm Ophelia from the Data {.tabset .tabset-pills}

```{r echo = FALSE, fig.cap = "Figure 2.1"}
ggplot(data, aes(month)) + geom_col(aes(y = wdsp, fill = wdsp)) +
  theme(axis.text.x = element_text(angle = 50, vjust = 1, hjust = 1, size = 12),
        axis.text.y=element_blank (), axis.ticks.y=element_blank ()) +
  ggtitle("Wind Speed per Month") +
  xlab("Month") + ylab("")
```
Figure 2.1 is a plot of the total wind speed per month with the blue lightening 
as the wind speed increases, there is a very light stripe occurring in October, 
showing that there was a time of very high wind speeds. It is seen here that for
the month of October the wind speed is the highest.

```{r}
plot1 <- ggplot(october_df, aes(as.factor(day))) +
  geom_line(aes(y = (msl)), color = "red", size = 2) +
  theme(axis.text.x = element_text(angle= 50, vjust = 1, hjust = 1, size = 12))+
  ggtitle("Mean Sea Level Pressure per Day") +
  xlab("Day") + ylab("Sea Level Pressure (hPa)") +
  labs(caption="Figure 2.2")

plot2 <- ggplot(october_df, aes(as.factor(day))) +
  geom_line(aes(y = (wdsp)), color = "blue", size = 2) +
  theme(axis.text.x = element_text(angle= 50, vjust = 1, hjust = 1, size = 12))+
  ggtitle("Wind Speed per Day") +
  xlab("Day") + ylab("Wind Speed (km/hr)") +
  labs(caption="Figure 2.3")

grid.arrange(plot1 , plot2)
```

Figure 2.2 shows that, on the 16th of October, there was an instance of 
extremely low pressure of 962.2 hPa. This matches with figure 2.3, which, on the
same day, had records of the highest wind speeds of the month at 109.268 km/hr,
again showing the relationship between pressure and wind speed.

```{r}
data %>% 
  slice_max(wdsp) %>% 
  select(station,county,wdsp,month,day,hour,date)
```
Here we can see that the highest wind speed occurred in Roches Point in Cork at
12:00pm.


### Storm Ophelia Analysis {.tabset .tabset-pills}

```{r}
plot1 <- ggplot(storm_ophelia, aes(as.factor(hour))) +
  geom_line(aes(y = (msl)), color = "red", size = 2) +
  theme(axis.text.x = element_text(angle= 50, vjust = 1, hjust = 1, size = 12))+
  ggtitle("Mean Sea Level Pressure per Hour") +
  xlab("Hour") + ylab("Sea Level Pressure (hPa)") +
  labs(caption="Figure 3.1")

plot2 <- ggplot(storm_ophelia, aes(as.factor(hour))) +
  geom_line(aes(y = (wdsp)), color = "blue", size = 2) +
  theme(axis.text.x = element_text(angle= 50, vjust = 1, hjust = 1, size = 12))+
  ggtitle("Wind Speed per Hour") +
  xlab("Hour") + ylab("Wind Speed (km/hr)") +
  labs(caption="Figure 3.2")

grid.arrange(plot1, plot2)
```

The lowest mean sea level pressure is seen in Figure 3.1 at a land station was 
observed at 962.2 hPa recorded at Valentia Observatory (County Kerry) on 16th of
October 2017 when the storm Ophelia occurred. Also, a wind speed of 
109.268 km/hr was recorded at Roches Point (County Cork) on the same day as per 
Figure 3.2 and the analysis performed.

### Removing Storm Ophelia Data {.tabset .tabset-pills}

```{r}
plot1 <- ggplot(not_storm_ophelia,
                aes(fill = factor(ifelse(month=="October","Highlighted","Normal")))) +
  geom_col(mapping = aes(x = as.factor(month), y = wdsp)) +
  ggtitle("Without Storm Ophelia") +
  xlab("Month") + ylab("Total Wind Speed (km/hr)") +
  coord_flip()+
  theme(plot.title = element_text(size = 10, face = "bold"))+     
  scale_fill_manual(name = "month", values=c("red","grey50")) +
  theme(legend.position="none") + 
  labs(caption="Figure 4.1")

plot2 <- ggplot(data,aes(fill = factor(ifelse(month=="October","Highlighted","Normal")))) +
  geom_col(mapping = aes(x = as.factor(month), y = wdsp)) +
  ggtitle("With Storm Ophelia",) +
  xlab("Month") + ylab("Total Wind Speed (km/hr)") +
  coord_flip()+
  theme(plot.title = element_text(size = 10, face = "bold"))+
  scale_fill_manual(name = "month", values=c("red","grey50")) +
  theme(legend.position="none") + 
  labs(caption="Figure 4.2")

grid.arrange(plot1, plot2, nrow=1, ncol=2)
```

This is a comparison of the total wind speed per month with Figure 4.1
containing the storm Ophelia data and Figure 4.2 with the storm 
Ophelia data removed. There is a decrease in total wind speed in 
October when the storm Ophelia data is removed, with the October total becoming
smaller than the March total.


## Conclusions {.tabset .tabset-pills}

So we have successfully completed the analyses of the aimsir17 dataset, we 
have analysed and compared rain, temperature, wind speed and mean sea level 
pressure and concluded that wind speed and mean sea level pressure were 
correlated.

While focusing on the relationship between wind speed and mean sea level 
pressure we found that the 16th of October had very high wind speed and low 
mean sea level pressure. Storm Ophelia occurred on this day, which is noticeable 
from this analysis.

Storm Ophelia was also analysed, we evaluated how high the wind speed was and 
how low the mean sea level pressure was hourly on that particular day, we then 
observed that the total wind speed was affected significantly by the data from
the day storm Ophelia occurred.

